% Copyright (C) 2014-2020 by Thomas Auzinger <thomas@auzinger.name>

\documentclass[draft,final]{vutinfth} % Remove option 'final' to obtain debug information.

% Load packages to allow in- and output of non-ASCII characters.
\usepackage{lmodern}        % Use an extension of the original Computer Modern font to minimize the use of bitmapped letters.
\usepackage[T1]{fontenc}    % Determines font encoding of the output. Font packages have to be included before this line.
\usepackage[utf8]{inputenc} % Determines encoding of the input. All input files have to use UTF8 encoding.

% Extended LaTeX functionality is enables by including packages with \usepackage{...}.
\usepackage{amsmath}    % Extended typesetting of mathematical expression.
\usepackage{amsthm}
\usepackage{listings}
\usepackage{amssymb}    % Provides a multitude of mathematical symbols.
\usepackage{mathtools}  % Further extensions of mathematical typesetting.
\usepackage{microtype}  % Small-scale typographic enhancements.
\usepackage[inline]{enumitem} % User control over the layout of lists (itemize, enumerate, description).
\usepackage{multirow}   % Allows table elements to span several rows.
\usepackage{booktabs}   % Improves the typesettings of tables.
\usepackage{subcaption} % Allows the use of subfigures and enables their referencing.
\usepackage[ruled,linesnumbered,algochapter]{algorithm2e} % Enables the writing of pseudo code.
\usepackage[usenames,dvipsnames,table]{xcolor} % Allows the definition and use of colors. This package has to be included before tikz.
\usepackage{nag}       % Issues warnings when best practices in writing LaTeX documents are violated.
\usepackage{todonotes} % Provides tooltip-like todo notes.
\usepackage{color}
\usepackage{float}
\usepackage{boldline}
\usepackage{hyperref}  % Enables cross linking in the electronic document version. This package has to be included second to last.
\usepackage[acronym,toc]{glossaries} % Enables the generation of glossaries and lists fo acronyms. This package has to be included last.

% Define convenience functions to use the author name and the thesis title in the PDF document properties.
\newcommand{\authorname}{Michael Langowski} % The author name without titles.
\newcommand{\thesistitle}{Evolog - Actions and Modularization in Lazy-Grounding Answer Set Programming} % The title of the thesis. The English version should be used, if it exists.

% Set PDF document properties
\hypersetup{
    pdfpagelayout   = TwoPageRight,           % How the document is shown in PDF viewers (optional).
    linkbordercolor = {Melon},                % The color of the borders of boxes around crosslinks (optional).
    pdfauthor       = {\authorname},          % The author's name in the document properties (optional).
    pdftitle        = {\thesistitle},         % The document's title in the document properties (optional).
    pdfsubject      = {Subject},              % The document's subject in the document properties (optional).
    pdfkeywords     = {a, list, of, keywords} % The document's keywords in the document properties (optional).
}

\setpnumwidth{2.5em}        % Avoid overfull hboxes in the table of contents (see memoir manual).
\setsecnumdepth{subsection} % Enumerate subsections.

\nonzeroparskip             % Create space between paragraphs (optional).
\setlength{\parindent}{0pt} % Remove paragraph identation (optional).

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\newtheorem{theorem}{Theorem}[section]

\newtheorem{example}{Example}[section]

\newtheorem{corollary}{Corollary}[section]

\definecolor{pblue}{rgb}{0.13,0.13,1}
\definecolor{pgreen}{rgb}{0,0.5,0}
\definecolor{pred}{rgb}{0.9,0,0}
\definecolor{pgrey}{rgb}{0.46,0.45,0.48}
\definecolor{ppurple}{rgb}{0.49,0.18,0.52}

\lstdefinestyle{code}{
	basicstyle=\ttfamily
}

\lstdefinestyle{asp-code}{
	basicstyle=\ttfamily,
	frame=single,
	numbers=left,
  stepnumber=1,
  breaklines=true,
  showstringspaces=false,
  breakatwhitespace=true,  
}

\lstdefinestyle{java}{language=Java,
  showspaces=false,
  showtabs=false,
  breaklines=true,
  showstringspaces=false,
  breakatwhitespace=true,
  commentstyle=\color{pgreen},
  keywordstyle=\color{pblue},
  stringstyle=\color{pred},
  basicstyle=\ttfamily,
  moredelim=[il][\textcolor{ppurple}]{$$},
  moredelim=[is][\textcolor{ppurple}]{\%\%}{\%\%}
}

% Set persons with 4 arguments:
%  {title before name}{name}{title after name}{gender}
%  where both titles are optional (i.e. can be given as empty brackets {}).
\setauthor{}{\authorname}{BSc.}{male}
\setadvisor{Prof. Dr.}{Thomas Eiter}{}{male}

% For bachelor and master theses:
\setfirstassistant{Dr.}{Antonius Weinzierl}{}{male}
%\setsecondassistant{Pretitle}{Forename Surname}{Posttitle}{male}
%\setthirdassistant{Pretitle}{Forename Surname}{Posttitle}{male}

% For dissertations:
%\setfirstreviewer{Pretitle}{Forename Surname}{Posttitle}{male}
%\setsecondreviewer{Pretitle}{Forename Surname}{Posttitle}{male}

% For dissertations at the PhD School and optionally for dissertations:
%\setsecondadvisor{Pretitle}{Forename Surname}{Posttitle}{male} % Comment to remove.

% Required data.
\setregnumber{01426581}
\setdate{09}{06}{2025} % Set date with 3 arguments: {day}{month}{year}.
\settitle{\thesistitle}{\thesistitle} % Sets English and German version of the title (both can be English or German). If your title contains commas, enclose it with additional curvy brackets (i.e., {{your title}}) or define it as a macro as done with \thesistitle.
\setsubtitle{}{} % Sets English and German version of the subtitle (both can be English or German).

% Select the thesis type: bachelor / master / doctor / phd-school.
% Bachelor:
%\setthesis{bachelor}
%
% Master:
\setthesis{master}
\setmasterdegree{dipl.} % dipl. / rer.nat. / rer.soc.oec. / master
%
% Doctor:
%\setthesis{doctor}
%\setdoctordegree{rer.soc.oec.}% rer.nat. / techn. / rer.soc.oec.
%
% Doctor at the PhD School
%\setthesis{phd-school} % Deactivate non-English title pages (see below)

% For bachelor and master:
\setcurriculum{Logic and Computation}{Logic and Computation} % Sets the English and German name of the curriculum.

\newcommand{\IDs}{\mathit{ID}}
\newcommand{\INTs}{\mathit{INT}}
\newcommand{\VARs}{\mathit{VAR}}

\newcommand{\NOT}{\mathit{not}}

\newcommand{\DEP}[1]{\succ_{d}^{#1}}

\newacronym{asp}{ASP}{Answer Set Programming}
\newacronym{cdnl}{CDNL}{Conflict-driven Nogood Learning}
\newacronym{cbp}{CBP}{Common Base Program}
\newacronym{dtd}{DTD}{Document Type Definition}
\newacronym{dom}{DOM}{Document Object Model}

\makeindex      % Use an optional index.
\makeglossaries % Use an optional glossary.
%\glstocfalse   % Remove the glossaries from the table of contents.

\begin{document}

\frontmatter % Switches to roman numbering.
% The structure of the thesis has to conform to the guidelines at
%  https://informatics.tuwien.ac.at/study-services

\addtitlepage{naustrian} % German title page (not for dissertations at the PhD School).
\addtitlepage{english} % English title page.
\addstatementpage

\begin{danksagung*}
Die vorliegende Abschlussarbeit ist das Ergbnis eines langen, oft zähen, und von vielen Ablenkungen - angefangen von den Anforderungen einer Vollzeit-Anstellung bis hin zu kleineren Ärgernisse wie einer Pandemie - geprägten Prozesses. Als ich die letzte Prüfung meines Studiums abschloss, war ich sicher, die Arbeit in wenigen Monaten fertigstellen zu können. Nun, mehr als ein Jahr später, ist es Zeit, allen Beteiligten für ihre Geduld und konstante Unterstützung zu danken. An erster Stelle steht hier Antonius Weinzierl, dessen Entwicklung von Alpha erst meine darauf aufbauende Arbeit ermöglichte, und der mir als Diskussionspartner und Advocatus Diavoli für die Formulierung und Abgrenzung meiner Vision stets zur Verfügung stand. Mindestens ebenso großer Dank gebührt meinem Betreuer Prof. Thomas Eiter, der in meinem langsamen Fortschritt stets den Fortschritt sah und hier größere Geduld bewies als ich selbst. Seinem stets offenen Ohr und bereitwillig gegebenem Feedback ist es zu verdanken, dass aus wenig zusammenhängenden Textschnipseln ein kohärenter Text wachsen konnte.

Mein herzlicher Dank für ihren moralischen Support durch alle Entstehungsstufen dieser Arbeit und ihren Glauben an mich und den Erfolg meines Vorhabens geht an Tanja Travnicek. Darüber hinaus danke ich meinen Freund*innen und meiner Familie für ihre Unterstützung während meines gesamten Studiums - es war eine lange Reise. \emph{Ich habe fertig.}
\end{danksagung*}

\begin{acknowledgements*}
This thesis is the result of a long and oftentimes arduous process characterized by distractions ranging from small things like a global pandemic to the demands of full-time employment. When I finished the last exam from my curriculum, I was convinced I'd finish this thesis within months. Now, more than a year later, it's time to thank all those who contributed to this success. First here is Antonius Weinzierl without whose pioneering work on the Alpha system this work wouldn't have been possible. He was also readily available as a discussion partner and sounding board in the early stages of the project and helped me greatly in formulating my vision. At least equally significant is the support I got from my advisor, Prof. Thomas Eiter, who, in my slow progress always saw progress first, and demonstrated more patience than I often had myself. His readily given Feedback and support is to thank for some vaguely related text snippets finally growing into a coherent text.

My heartfelt thanks for her moral support and confidence in me and the eventual success of my project goes to Tanja Travnicek. In addition, I want to thank my friends and family for their continuing support through all stages of my studies - It has been a long road. \emph{I'm done.}
\end{acknowledgements*}

\begin{kurzfassung}
  \emph{Answer Set Programming (ASP)} ist ein seit Jahrzehnten etablierter Formalismus in der Logikprogrammierung, der insbesondere bei der Modellierung und Lösung NP-schwerer Planungsprobleme Anwendung findet. Effiziente Solver für ASP sind seit über 20 Jahren verfügbar. Aktuelle Trends in ASP befassen sich zunehmend mit der Interaktion mit externen Daten - sowohl zum Lesen und Verarbeiten selbiger, angefangen von Stream Reasoning bis hin zu Interaktionen mit Machine-Learning-Software - als auch zum Ausführen von Aktionen, d.h. der Entwicklung von ASP-Programmen mit Seiteneffekten (side-effects). Das primäre Ziel der vorliegenden Arbeit ist es, eine neue Methode zur Modellierung von Regeln mit Seiteneffekten in ASP Programmen zu entwickeln, die sich leicht in einem ASP-Solver wie dem von Weinzierl et al. entwickelten \emph{Lazy-Grounding}-Solver \emph{Alpha} implementieren lässt. Ein weiteres Ziel ist die Entwicklung eines leichtgewichtigen Konzeptes zur Programmmodularisierung, das die zuvor erwähnte Unterstützung von Aktionen ergänzt. Die Notwendigkeit hierfür begründet sich aus der Tatsache, das Modellieren von Seiteneffekten in einer vollständig deklarativen Sprache zwangsläufig einige Einschränkungen struktureller Natur für Programme mit Aktionen mit sich bringt. Die mit dem Modularisierungskonzept gegebene Möglichkeit, seiteneffekt-freie Unterprogramme als eigenständige Einheiten auszuführen, sollte diese Einschränkungen im Praxiseinsatz ausgleichen. Die resultierende erweiterte Answer-Set-Programming-Sprache wird \emph{Evolog} genannt.
  Die formale Semantik von Aktionen wird bezüglich eines \emph{Frame}, d.h. eines Modells der Seiteneffekte einer Aktion in der Außenwelt in Form einer Interpretationsfunktion für \emph{Action-Atome}, modelliert. Basierend auf diesem Modell der Außenwelt werden Kriterien definiert, nach denen ein Evolog-Programm (semantisch) gültig ist. Schließlich wird das Konzept eines \emph{Evolog-Modells} definiert, d.h. eines Stable-Models eines Evolog-Programms, das konsistent bezüglich einem gegebenen Frame ist. Darüber hinaus wird gezeigt, dass Evolog eine \emph{konservative} Erweiterung der üblichen Stable-Model-Semantik ist, insofern als jedes reguläre ASP-Programm auch ein gültiges Evolog-Programm in allen Frames ist.
  Die Modularisierung wird basierend auf Alpha's bestehendem Konzept eines \emph{externen Atoms} modelliert, d.h. eines Atoms, das eine seiteneffektfreie Funktion abbildet, die in einer beliebigen Programmiersprache (in diesem Fall Java) implementiert ist. Modul-Atome sind definiert als externe Atome, deren Interpretationsfunktion darin besteht, einen ASP-Solver mit einem fixen Programm (der Modul-Implementierung) und den Eingabetermen des externen Atoms als zusätzlichem Input aufzurufen. Um das Schreiben von Code, der Module benutzt, effizienter zu gestalten, wird das (rein syntaktische) Konzept eines \emph{Listen-Terms} eingeführt.
  Um die Eignung von Evolog als tatsächliche Programmiersprache zu demonstrieren, wird eine einfache Kommandozeilenapplikation vorgestellt, die Graphen aus XML-Dateien einliest, 3-Colorings dieser Graphen berechnet und die berechneten Colorings in XML-Dateien serialisiert. Die Applikation wird sowohl in Bezug auf die Bedienbarkeit von Evolog als Implementierungssprache als auch auf die Ausführungsleistung diskutiert. Basierend auf der Demonstration anhand der XML-3-Coloring-App werden mögliche Erweiterungen der Sprache, sowie zukünftige Wege zur Steigerung der Solver-Performance angesprochen.
\end{kurzfassung}

\begin{abstract}
For decades, \emph{Answer Set Programming (ASP)} has been an established formalism in logic programming, especially used for NP-hard planning problems, with efficient solvers available for more than 20 years. Current trends in ASP increasingly deal with interaction with external data, both for reading and processing (such as in stream reasoning or even interacting with machine learning software) as well as performing actions, i.e. influencing the state of the outside world. This thesis' primary concern is exploring a new way of representing actions in a manner suitable for easy implementation in a \emph{lazy-grounding} ASP solver such as Alpha by Weinzierl et al. Secondary to that, the other goal of this work is to specify and implement a light-weight program modularization concept to go along with the aforementioned action support. The inspiration here is that, since inevitably modelling side-effects in a fully declarative language forces us to impose some restrictions on programs with actions, being able to run action-free subprograms as encapsulated units should alleviate any such strictures in software development. The resulting extended answer set programming language is called \emph{Evolog}.
We define the formal semantics of actions with regards to a \emph{frame}, i.e. a formalization of the effects of an action in the outside world, that is modelled as an interpretation function for \emph{action atoms}. Based on this model of the outside world, the notion of what constitutes a (semantically) valid Evolog program is defined. Finally, we define an \emph{Evolog Model}, i.e. a stable model of an Evolog program that is consistent with a given frame. We further show that Evolog is a \emph{conservative} extension to traditional ASP in the sense that every regular ASP program is also a valid Evolog program in all frames.
Modularization is modelled around Alpha's existing concept of an \emph{external atom}, i.e. an atom that models a side-effect free function which is implemented in an arbitrary programming language (Java in this case). We define module atoms as external atoms whose interpretation function is given as calling an ASP solver with a fixed program (the module body) and the input terms of the external atom as additional input. To aid in efficiently writing code around modules, the purely syntactical concept of a \emph{list term} is introduced.
In order to demonstrat Evolog's suitability as an actual application programming language, a simple command-line application for reading graphs from XML files, calculating 3-colorings of said graphs and serializing calculated colorings to XML files, is showcased and discussed in terms of both ease of development, as well as evaluation performance. Based on our demonstration of the XML-3-coloring application, we formulate possible refinements to the language as well as future paths for improvement in solver performance.
\end{abstract}

% Select the language of the thesis, e.g., english or naustrian.
\selectlanguage{english}

% Add a table of contents (toc).
\tableofcontents % Starred version, i.e., \tableofcontents*, removes the self-entry.

% Switch to arabic numbering and start the enumeration of chapters in the table of content.
\mainmatter

\chapter{Introduction}
\input{intro.tex} % A short introduction to LaTeX.

\chapter{Preliminaries}
\input{preliminaries.tex}

\chapter{The Evolog Language}
\label{chap:language}
\input{language.tex}

\chapter{Evolog Reference Implementation}
\label{chap:implementation}
\input{implementation.tex}

\chapter{Experimental Application}
\label{chap:application-experiment}
\input{application-experiment.tex}

\chapter{Discussion}
\label{chap:discussion}
\input{discussion.tex}

\chapter{Conclusion}
\label{chap:conclusion}
\input{conclusion.tex}

\appendix
\input{appendix/appendix.tex}

\backmatter

% Use an optional list of figures.
\listoffigures % Starred version, i.e., \listoffigures*, removes the toc entry.

% Use an optional list of tables.
\cleardoublepage % Start list of tables on the next empty right hand page.
\listoftables % Starred version, i.e., \listoftables*, removes the toc entry.

% Use an optional list of alogrithms.
\listofalgorithms
\addcontentsline{toc}{chapter}{List of Algorithms}

% Add an index.
\printindex

% Add a glossary.
\printglossaries

% Add a bibliography.
\bibliographystyle{alpha}
\bibliography{evolog}

\end{document}